agdarsec -- Total Parser Combinators

### 2017-10-13 doc/paper/main.tex
[ more ] start of parsing, totally

Parser combinators have made functional languages such as Haskell
shine: they are a prime example of the advantages Embedded Domain
Specific Languages provide the end user. She not only has access
to a library of powerful composable abstractions but she is also
able to rely on the existing tooling and libraries of the host
language, she gets feedback from the static analysis built in the
compiler (e.g. type and coverage checking) and can exploit from
the expressivity of the host language to write generic parsers
thanks to polymorphism and higher order functions.

@@

\paragraph{Parser Type}

@@

It is naturally possible to run such a parser and try to extract
a value from a successful parse. Opinions may vary on whether a
run with leftover characters can be considered successful or not.
We decide against it. This is not crucial as both styles can be
mutually emulated by either providing an 'end of file' parser
guaranteeing that only runs with no leftovers are successful or
by extending a grammar with a dummy case consuming the rest of
the input string.

@@

% insert alternative instance

% interesting use case? replicateM?

@@

@@

\section{The Issue with Haskell's Parser Types}

@@

\paragraph{Remark: Record Wrapper} Instead of defining \BOX{} as a
function like the other combinators, we wrap the corresponding
function space in a record type. This prevents normalisation from
unfolding the combinator too eagerly and makes types more readable
during interactive development.

\paragraph{Remark: Irrelevance} The argument stating that \AB{m}
is strictly smaller than \AB{n} is preceded by a dot. In Agda,
it means that this value is irrelevant and can be erased by the
compiler. In Coq, we would define the relation \AF{\_<\_} in
\type{Prop} to achieve the same.

@@

\paragraph{Remark: Copatterns} The definition of \AF{map} uses
the \BOX{} field named \ARF{call} on the \emph{left hand side}.
This is a copattern\todo{cite}, meaning that we explain how
the definition is \emph{observed} (via \ARF{call}) rather than
\emph{constructed} (via \AIC{mkBox}).

@@

\paragraph{Remark: Counit} The careful reader will have noticed
that this is not quite the extract we would expect from a comonad:
for a counit, we would need a natural transformation between
\BOX{} \AB{A} and \AB{A} i.e. a function of type
\AF{[} \BOX{} \AB{A} \AF{⟶} \AB{A} \AF{]}. We will not be able
to define such a function: \BOX{} \AB{A} \AN{0} is isomorphic to
the unit type so we would have to generate an \AB{A} \AN{0} out
of thin air.

\medskip

@@

\paragraph{Remark: Identifiers in Agda} In Agda, any space-free
string which is not a reserved keyword is a valid identifier. As
a consequence we can pick suggestive names such as \AB{m<n} for
a proof that $m < n$.

\agrab{Induction/Nat/Strong}{app}
\sagrab{Induction/Nat/Strong}{fix}

\paragraph{Remark: Generalisation} The type of \AF{fix} corresponds
exactly to strong induction for the natural numbers. A similar \BOX{}
type constructor can be defined for any induction principle relying
on an accessibility predicate. Which means that a library's types
can be cleaned up by using these combinators in any situation where
one had to give up structural induction for a more powerful alternative.

@@

>>>>

Parser combinators have made functional languages such as Haskell
shine. They are a prime example of the advantages Embedded Domain
Specific Languages provide the end user. She not only has access
to a library of powerful and composable abstractions but she is also
able to rely on the host language's existing tooling and libraries.
She also gets feedback from the static analyses built in the compiler
(e.g. type and coverage checking) and can exploit the expressivity
of the host language to write generic parsers thanks to polymorphism
and higher order functions.

However she only gets the guarantees the host language is willing
to give. In non-total programming languages such as Haskell this
means she will not be prevented from writing parsers which will
unexpectedly fail on some (or even all!) inputs. Handling a
left-recursive grammar is perhaps the most iconic pitfall leading
beginners to their doom: a parser never making any progress.

We start with a primer on parser combinators and follow up with the
definition of a broken parser which is silently accepted by Haskell.
We then move on to Agda~\todo{cite} and introduce combinators to
define functions by well-founded recursion. This allows us to define
a more informative notion of parser and give more precise types to
the combinators commonly used. We then demonstrate that broken parsers
such as the one presented earlier are rejected whilst typical example
can be ported with minimal modifications.

@@

\paragraph{Parser Type} Let us start by reminding ourselves what
a parser is. Although we will eventually move to a more generic
type, Fritz Ruehr's rhyme gives us the \emph{essence} of parsers:

\begin{verse}
  A Parser for Things\\
  is a function from Strings\\
  to Lists of Pairs\\
  of Things and Strings!
\end{verse}

This stanza directly translates to the following Haskell type (we
use a \textbf{newtype} wrapper to have cleaner error messages):

@@

It is naturally possible to run such a parser and try to extract
a value from a successful parse. Opinions may vary on whether a
run with leftover characters can be considered successful or not.
We decide against it.\todo{ambiguity} This is not crucial as both styles can be
mutually emulated by either providing an 'end of file' parser
guaranteeing that only runs with no leftovers are successful or
by extending a grammar with a dummy case consuming the rest of
the input string.

@@

% insert alternative instance
% interesting use case? replicateM?

@@

The two parsers \AB{some} and \AB{many} turn a parser for elements
into ones for non-empty and potentially empty lists of such elements
respectively. They concisely showcase the power of mutual recursion,
higher-order functions and the \type{Functor}, \type{Applicative},
and \type{Alternative} structure.

@@

\section{The Issue with Haskell's Parser Types}\label{sec:haskellproblem}

@@

\remark{Record Wrapper} Instead of defining \BOX{} as a
function like the other combinators, we wrap the corresponding
function space in a record type. This prevents normalisation from
unfolding the combinator too eagerly and makes types more readable
during interactive development.

\remark{Irrelevance} The argument stating that \AB{m}
is strictly smaller than \AB{n} is preceded by a dot. In Agda,
it means that this value is irrelevant and can be erased by the
compiler. In Coq, we would define the relation \AF{\_<\_} in
\type{Prop} to achieve the same.

@@

\remark{Copatterns} The definition of \AF{map} uses
the \BOX{} field named \ARF{call} on the \emph{left hand side}.
This is a copattern~\cite{abel2013copatterns}, meaning that we
explain how the definition is \emph{observed} (via \ARF{call})
rather than \emph{constructed} (via \AIC{mkBox}).

@@

\remark{Counit} The careful reader will have noticed
that this is not quite the extract we would expect from a comonad:
for a counit, we would need a natural transformation between
\BOX{} \AB{A} and \AB{A} i.e. a function of type
\AF{[} \BOX{} \AB{A} \AF{⟶} \AB{A} \AF{]}. We will not be able
to define such a function: \BOX{} \AB{A} \AN{0} is isomorphic to
the unit type so we would have to generate an \AB{A} \AN{0} out
of thin air.

@@

\remark{Identifiers in Agda} Any space-free string which
is not a reserved keyword is a valid identifier. As a consequence
we can pick suggestive names such as \AB{m<n} for a proof that $m < n$.

Exploring further the structure of the functor \BOX{}, we can observe
that just like it is not quite a comonad, it is not quite an applicative
functor. Indeed we can only define \AB{pure}, a natural transformation
of type \AF{[} \AB{A} \AF{⟶} \BOX{} \AB{A} \AF{]}, for the types \AB{A}
that are downwards closed. Providing the user with \AB{app} is however
possible:

\agrab{Induction/Nat/Strong}{app}

Finally, we can reach what will serve as the backbone of our parser
definitions: a safe, total fixpoint combinator. It differs from the
traditional $Y$ combinator in that all the recursive calls have to be
guarded.

\sagrab{Induction/Nat/Strong}{fix}

\remark{Generalisation} If we were to unfold all the type-level
combinators and record wrappers, the type of \AF{fix} would correspond
exactly to strong induction for the natural numbers. A similar \BOX{}
type constructor can be defined for any induction principle relying
on an accessibility predicate. Which means that a library's types
can be cleaned up by using these combinators in any situation where
one had to give up structural induction for a more powerful alternative.

@@

As already highlighted in Section~\ref{sec:haskellproblem}, \AB{some}
and \AB{many} can yield diverging computations if the parser they are
given as an argument succeeds on the empty string. To avoid any such
issue, we adopt a radical solution: for a parser's run to be considered
successful, it must have consumed some of its input.

\agrab{Text/Parser/Success}{success}

\agrab{Text/Parser/Combinators}{parser}
